A significance test in forward stepwise model selection

### 2013-12-01 paper/paper.tex
Nothing to see here
Adding to fwd stepwise section of paper

We allow forward stepwise selection to add groups of variables in each
step, not only in the case of binary encoding for categorical
variables but also for any grouping purpose. For example, groups of
variables may be pre-designated factors such as expression
measurements for all genes in a single regulatory pathway. An entire
group is included or excluded together from the final model. For
consistency we will use $g, h$ as indices rather than the usual $i, j$
throughout. Since single variables can be considered groups of size 1,
our general exposition includes non-grouped situations as a special case.

Label the outcome variable of $n$ i.i.d. measurements $y \in \real^n$. Let an integer $G \geq 1$ 
be the number of groups. For each $1 \leq g \leq G$ the design matrix encoding the
$g$th group is the $n \times p_g$ matrix denoted $X_g$, where $p_g$ is
the number of individual variables in group $g$. Define $p = \sum_{g=1}^G
p_g$ be the total number of individual variables, so $p = G$ in the
case where all groups have size 1. Let $X$ be the matrix constructed
by column-binding the $X_g$, that is  
%Throughout most of this paper we assume orthogonality within groups, that is
%$X_g^T X_g = I_{p_g \times p_g}$ for all $g$. ( \todo Take this out if
%we don't actually need it).  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation*}
X = \begin{pmatrix} X_1 & X_2 & \cdots & X_G  \end{pmatrix}
\end{equation*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
With each group we associate the $p_g \times 1$ coefficient vector
$\beta_g$, and write $\beta$ for the $p \times 1$ vector constructed
by stacking all of the $\beta_g$ in order.  Finally, our model for the
response is
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gmodel}
y & = X \beta + \sigma \epsilon \\
   & = \sum_{g=1}^G X_g \beta_g + \sigma \epsilon
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $\epsilon$ is noise. Unless otherwise specified we assume
i.i.d. Gaussian noise $\epsilon \sim N(0, I_{n \times n})$ and that
$\sigma$ is known.

@@

Before we describe the forward stepwise procedure we require one last
ingredient. To each group of variables we assign a weight $w_g$. These
weights act like penalties or costs, so increasing $w_g$ makes it
more difficult for the group $X_g$ to enter the model. The
modeler can choose weights arbitrarily, but we will only use one
particular choice, based on $p_g$, that we discuss later. With this we
are ready to describe the forward stepwise procedure, given in Algorithm
\ref{algo:fs}. Note that to prevent saturation, and to enable our p-value
computations, \textit{max.steps} should be at most $\min (n, G) - 1$.

\begin{algorithm}
\DontPrintSemicolon
\KwData{An $n$ vector $y$ and $n \times p$ matrix $X$ of $G$ grouped variables}
\KwResult{Active set $A$ of variable groups included in the model}
\SetKwFunction{lsfitResidual}{lsfitResidual}
\caption{Forward stepwise procedure with groups and weights}
\BlankLine
$A \gets \emptyset$\;
$A^c \gets \{ 1, \ldots, G \}$\;
$r_0 \gets y$\;
\For{$step \leftarrow 1$ \KwTo $max.steps$}{
  $g^* \gets \argmax_{g \in A^c} \norm{X_g^T r_{g-1}}_2 / w_g$\;
  $A \gets A \cup \{ g^* \}$\;
  $A^c \gets A^c \backslash \{ g^* \}$\;
  $r_g \gets $\lsfitResidual{$r_{g-1}, X_{g^*}$}\;
}
\KwRet{$A$}
\label{algo:fs}
\end{algorithm}

@@

So far we have left open the question of choosing among the models in
the forward stepwise sequence, i.e. when to stop stepping
forward. Some classical approaches can be posed as optimization
criteria and stop at the step minimizing
\begin{equation}
\begin{aligned}
\label{eq:subsetregress}
\frac{1}{2} \| y - X \beta_s \|_2^2 + \lambda \pen(\beta_s)
\end{aligned}
\end{equation}
where we have written $\{ \beta_s : s = 1, \ldots, max.steps \}$ as
the sequence of models outputted by forward stepwise. The function
$\pen(\beta)$ is a penalty on model complexity usually taken to be the
number of nonzero entries of $\beta$. Proposals for $\lambda$ include
2 ($C_p$ of \cite{CP}, AIC of \cite{AIC}), $\log(n)$ (BIC of \cite{BIC}), and
$2\log(p)$ (RIC of \cite{RIC}). In the current paper we provide
accurate p-values when adding the first noise variable group in the forward
stepwise sequence, and it is natural to consider using these p-values
to choose a model. \cite{sequential:fdr} examined some stopping rules
using the asymptotic p-values of \cite{significance:lasso} and showed
their stopping rules control false discovery rate--the expected
proportion of noise variables among variables declared significant
\citep{fdr}. We explore this further in Section \ref{sec:simulations}.

@@

\begin{figure}
\begin{center}
\subfigure[Gaussian design]{
\label{fig:smallgauss}
\includegraphics[width=0.5\textwidth]{../figs/small_sim_gaussian_design.pdf}}
\hspace{-15pt}
\subfigure[Categorical design]{
\label{fig:smallcat}
\includegraphics[width=0.5\textwidth]{../figs/small_sim_categorical_design.pdf}}
\caption{\small \it The left panel shows results of a simulation with
  an independent Gaussian design matrix and a signal vector having
  three groups with non-zero coefficients of sizes one, two, and
  three, and seven groups of noise variables with sizes varying from
  one to ten, for a total of 30 variables. For the right panel we
  repeated this setup, but group sizes corresponded to levels of
  categorical variables. The groups of size one were increased to two,
so there are 10 categorical variables with a total of 32 levels.}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfigure[Large Gaussian lasso]{
\label{fig:largelasso}
\includegraphics[width=0.5\textwidth]{../figs/large_sim_gaussian_lasso.pdf}}
\hspace{-15pt}
\subfigure[Large Gaussian group lasso]{
\label{fig:largegrplasso}
\includegraphics[width=0.5\textwidth]{../figs/large_sim_gaussian_design.pdf}}
\caption{\small \it The left panel shows results of a simulation with
  an independent Gaussian design matrix and a signal vector having
  three groups with non-zero coefficients of sizes one, two, and
  three, and seven groups of noise variables with sizes varying from
  one to ten, for a total of 30 variables. For the right panel we
  repeated this setup, but group sizes corresponded to levels of
  categorical variables. The groups of size one were increased to two,
so there are 10 categorical variables with a total of 32 levels.}
\end{center}
\end{figure}

>>>>

We allow forward stepwise selection to add groups of variables in each
step, not only in the case of binary encoding for categorical
variables but also for any grouping purpose. For example, groups of
variables may be pre-designated factors such as expression
measurements for all genes in a single functional pathway. An entire
group is included or excluded together from the final model. For
consistency we will use $g, h$ as indices rather than the usual $i, j$
throughout. Since single variables can be considered groups of size 1,
our general exposition includes non-grouped situations as a special case.

Label the outcome variable of $n$ i.i.d. measurements $y \in \real^n$. Let an integer $G \geq 1$ 
be the number of groups. For each $1 \leq g \leq G$ the design matrix encoding the
$g$th group is the $n \times p_g$ matrix denoted $X_g$, where $p_g$ is
the number of individual variables in group $g$. Define $p = \sum_{g=1}^G
p_g$ the total number of individual variables, so $p = G$ in the
case where all groups have size 1. Let $X$ be the matrix constructed
by column-binding the $X_g$, that is  
%Throughout most of this paper we assume orthogonality within groups, that is
%$X_g^T X_g = I_{p_g \times p_g}$ for all $g$. ( \todo Take this out if
%we don't actually need it).  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation*}
X = \begin{pmatrix} X_1 & X_2 & \cdots & X_G  \end{pmatrix}
\end{equation*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
With each group we associate the $p_g \times 1$ coefficient vector
$\beta_g$, and write $\beta$ for the $p \times 1$ vector constructed
by stacking all of the $\beta_g$ in order.  Finally, our model for the
response is
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gmodel}
y & = X \beta + \sigma \epsilon \\
   & = \sum_{g=1}^G X_g \beta_g + \sigma \epsilon
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $\epsilon$ is noise. Unless otherwise specified we assume
i.i.d. Gaussian noise $\epsilon \sim N(0, I_{n \times n})$ and that
$\sigma$ is known.

@@

Before we describe the forward stepwise procedure we require one last
ingredient. To each group of variables we assign a weight $w_g$. These
weights act like penalties or costs, so increasing $w_g$ makes it
more difficult for the group $X_g$ to enter the model. The
modeler can choose weights arbitrarily, but we will only use one
particular choice, based on $p_g$, that we discuss later. With this we
are ready to describe the forward stepwise procedure, given in Algorithm
\ref{algo:fs}. Note that to enable our p-value computations,
\textit{max.steps} should be at most $\min (n, G) - 1$. In our
implementation we treat the active set $A$ as an ordered set to easily
track the order of variables entering the model.

\begin{algorithm}
\DontPrintSemicolon
\KwData{An $n$ vector $y$ and $n \times p$ matrix $X$ of $G$ grouped variables}
\KwResult{Active set $A$ of variable groups included in the model}
\SetKwFunction{lsfitResidual}{lsfitResidual}
\caption{Forward stepwise procedure with groups and weights}
\BlankLine
$A \gets \emptyset$\;
$A^c \gets \{ 1, \ldots, G \}$\;
$r_0 \gets y$\;
\For{$step \leftarrow 1$ \KwTo $max.steps$}{
  $g^* \gets \argmax_{g \in A^c} \norm{X_g^T r_{g-1}}_2 / w_g$\;
  $A \gets A \cup \{ g^* \}$\;
  $A^c \gets A^c \backslash \{ g^* \}$\;
  \For{$g \in A^c$}{
    $X_g \gets (I - X_{g^*}X_{g^*}^\dagger) X_g$;
  }
  $r_g \gets $\lsfitResidual{$r_{g-1}, X_{g^*}$}\;
}
\KwRet{$A$}
\label{algo:fs}
\end{algorithm}

@@

So far we have left open the question of choosing among the models in
the forward stepwise sequence, i.e. when to stop stepping
forward. Some classical approaches can be posed as optimization
criteria which stop at the step minimizing
\begin{equation}
\begin{aligned}
\label{eq:subsetregress}
\frac{1}{2} \| y - X \beta_s \|_2^2 + \lambda \pen(\beta_s)
\end{aligned}
\end{equation}
where we have written $\{ \beta_s : s = 1, \ldots, max.steps \}$ as
the sequence of models outputted by forward stepwise. The function
$\pen(\beta)$ is a penalty on model complexity usually taken to be the
number of nonzero entries of $\beta$. Proposals for $\lambda$ include
2 ($C_p$ of \cite{CP}, AIC of \cite{AIC}), $\log(n)$ (BIC of \cite{BIC}), and
$2\log(p)$ (RIC of \cite{RIC}). In the current paper we provide
accurate p-values when adding the first noise variable group in the forward
stepwise sequence, and it is natural to consider using these p-values
to choose a model. \cite{sequential:fdr} examined some stopping rules
using the asymptotic p-values of \cite{significance:lasso} and showed
their stopping rules control false discovery rate--the expected
proportion of noise variables among variables declared significant
\citep{fdr}. We explore this further in Section \ref{sec:simulations}.

Although forward stepwise is a greedy algorithm producing a
potentially sub-optimal sequence of models, under favorable conditions
it can still perform well. There is a parallel literature to the
compressed sensing literature \textit{(cite some here)} with theorems
stating that forward stepwise can exactly select the true model under
some stringent conditions involving quantities like the sparsity of
the true model and the coherence of the design matrix. We do not
pursue this any further because we find empirically that forward
stepwise can work well in much less restrictive situations than those
guaranteed by theorems.

To explore the performance of forward stepwise empirically we
conducted simulations with results shown in ~\ref{fig:fwdstepsims}

\begin{figure}
\begin{center}
\subfigure[Gaussian design]{
%\label{fig:smallgauss}
\includegraphics[width=0.5\textwidth]{../figs/fwdstep/gaussian_p2n_n100_lower1p4_upper1p6.pdf}}
\hspace{-15pt}
\subfigure[Categorical design]{
%\label{fig:smallcat}
\includegraphics[width=0.5\textwidth]{../figs/fwdstep/gaussian_p2n_n100_lower1p4_upper1p6_corr0p2.pdf}}
\caption{\small \it The left panel shows results of a simulation with
  an independent Gaussian design matrix and a signal vector having
  three groups with non-zero coefficients of sizes one, two, and
  three, and seven groups of noise variables with sizes varying from
  one to ten, for a total of 30 variables. For the right panel we
  repeated this setup, but group sizes corresponded to levels of
  categorical variables. The groups of size one were increased to two,
so there are 10 categorical variables with a total of 32 levels.}
\label{fig:fwdstepsim}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfigure[Gaussian design]{
\label{fig:smallgauss}
\includegraphics[width=0.5\textwidth]{../figs/gaussian_size1_n100_p200_g200_k10_lower1p8_upper1p9.pdf}}
\hspace{-15pt}
\subfigure[Categorical design]{
\label{fig:smallcat}
\includegraphics[width=0.5\textwidth]{../figs/gaussian_size1_n100_p200_g200_k10_lower1p8_upper1p9_corr0p1.pdf}}
\caption{\small \it The left panel shows results of a simulation with
  an independent Gaussian design matrix and a signal vector having
  three groups with non-zero coefficients of sizes one, two, and
  three, and seven groups of noise variables with sizes varying from
  one to ten, for a total of 30 variables. For the right panel we
  repeated this setup, but group sizes corresponded to levels of
  categorical variables. The groups of size one were increased to two,
so there are 10 categorical variables with a total of 32 levels.}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfigure[Large Gaussian lasso]{
\label{fig:largelasso}
\includegraphics[width=0.5\textwidth]{../figs/gaussian_size5-10_n100_p200_g35_k10_lower1p8_upper1p9.pdf}}
\hspace{-15pt}
\subfigure[Large Gaussian group lasso]{
\label{fig:largegrplasso}
\includegraphics[width=0.5\textwidth]{../figs/gaussian_size5-10_n100_p200_g35_k10_lower1p8_upper1p9_corr0p1.pdf}}
\caption{\small \it The left panel shows results of a simulation with
  an independent Gaussian design matrix and a signal vector having
  three groups with non-zero coefficients of sizes one, two, and
  three, and seven groups of noise variables with sizes varying from
  one to ten, for a total of 30 variables. For the right panel we
  repeated this setup, but group sizes corresponded to levels of
  categorical variables. The groups of size one were increased to two,
so there are 10 categorical variables with a total of 32 levels.}
\end{center}
\end{figure}
